---
title: "Extreme Value Statistics - project"
author:
  - Tereza Ehnová
  - Maciej Chylak
output: html_document
date: '2023-03-15'
---


```{r, include=FALSE}
library(dplyr)
library(extRemes)
library(ggplot2)
library(ismev)
library(evd)
library(reshape)
library('ROOPSD')
library('VineCopula')
library(sp)
```

### Loading data and preprocessing

```{r}
load("G8_Rainfall.RData")

colnames(dataset) <- c("loc_1", "loc_2", "loc_3", "loc_4", "loc_5")
head(dataset)
```
```{r}
rainfall_g8 <- as.data.frame(dataset)
rainfall_g8["day"] <- as.Date(as.character(dates))
rainfall_g8["month_n"] <- format(rainfall_g8["day"], "%m")
rainfall_g8["year_n"] <- format(rainfall_g8["day"], "%Y")
rainfall_g8["year_month_n"] <- format(rainfall_g8["day"], "%Y%m")
rainfall_g8["weekday"] <- factor(weekdays(rainfall_g8$day, abbreviate=TRUE), levels=c("Mon","Tue","Wed","Thu", "Fri", "Sat", "Sun"), ordered=TRUE)

rainfall_g8$month_n <- as.numeric(rainfall_g8$month_n)
rainfall_g8$year_n <- as.numeric(rainfall_g8$year_n)
rainfall_g8$year_month_n <- as.numeric(rainfall_g8$year_month_n)

rainfall_g8$month_c <- as.character(rainfall_g8$month_n)
rainfall_g8$year_c <- as.character(rainfall_g8$year_n)
rainfall_g8$year_month_c <- as.character(rainfall_g8$year_month_n)

rainfall_g8_up <- melt(rainfall_g8, id=c("day", "month_n", "year_n", "year_month_n", "month_c", "year_c", "year_month_c", "weekday"))
rainfall_g8_up$value <- as.numeric(rainfall_g8_up$value)

colnames(rainfall_g8_up) <- c("day", "month_n", "year_n", "year_month_n", "month_c", "year_c", "year_month_c", "weekday", "location", "value")
```

```{r}
summary(rainfall_g8[, c("loc_1", "loc_2", "loc_3", "loc_4", "loc_5")])
```
### Correlation of rainfall in weather stations

```{r, warning=FALSE}
df_cor <- melt(round(cor(subset(rainfall_g8, select=c(loc_1, loc_2, loc_3, loc_4, loc_5))), 2))

ggplot(df_cor, aes(x=X2, y=X1, fill=value)) +
  geom_tile() +
  scale_fill_gradient2(low="blue", high="red", mid="white", midpoint=0, limit=c(-1, 1), name="correlation") +
  labs(title="Correlation matrix of rainfall in each location" , y="", x="") + 
  theme(plot.title=element_text(color="black", size=16, face="bold.italic", hjust="0.5")) +
  geom_text(aes(X2, X1, label=value))
```
Recorded rainfall at weather stations is highly correlated with each other. These station are probably not far away -> Tobler's First Law of Geography

```{r}
head(rainfall_g8_up)
```

### Distribution of rainfall data

```{r include=FALSE}
rainfall_g8_gyears <- rainfall_g8_up %>% 
  group_by(year_c, year_n, location) %>%
  summarise(
    total_rainfall = sum(value),
    mean_rainfall = mean(value),
    max_rainfall = max(value),
  )

rainfall_g8_gweekday <- rainfall_g8_up %>% 
  group_by(weekday, location) %>%
  summarise(
    total_rainfall = sum(value),
    mean_rainfall = mean(value),
    max_rainfall = max(value),
  )

rainfall_g8_gmonth <- rainfall_g8_up %>% 
  group_by(month_c, month_n, location) %>%
  summarise(
    total_rainfall = sum(value),
    mean_rainfall = mean(value),
    max_rainfall = max(value),
  )


rainfall_g8_gyear_month <- rainfall_g8_up %>% 
  group_by(year_month_c, year_month_n, location) %>%
  summarise(
    total_rainfall = sum(value),
    mean_rainfall = mean(value),
    max_rainfall = max(value),
  )

```

```{r}
ggplot(data=rainfall_g8_gyears, aes(x=year_n, y=mean_rainfall, color=location)) + 
  geom_line()  +
  labs(title="Timeseries of average of rainfall in each location" , y="Rainfall", x="Years", legend="Location") +    theme(plot.title=element_text(color="black", size=16, face="bold.italic", hjust="0.5"))
```

```{r}
ggplot(data=rainfall_g8_gyears, aes(x=year_n, y=max_rainfall, color=location)) + 
  geom_line()  +
  labs(title="Timeseries of max of rainfall in each location" , y="Rainfall", x="Years", legend="Location") +    theme(plot.title=element_text(color="black", size=16, face="bold.italic", hjust="0.5"))
```

```{r}
ggplot(data=rainfall_g8_gyears, aes(x=year_n, y=total_rainfall, color=location)) + 
  geom_line()  +
  labs(title="Timeseries of total of rainfall in each location" , y="Rainfall", x="Years", legend="Location") +    theme(plot.title=element_text(color="black", size=16, face="bold.italic", hjust="0.5"))
```
Rainfall data seems to be stationary. We don't need to perform any related transformations

Also rainfall data is mixing. Value at widely-separated times are asymptotically independent (due to the nature of weather data).

```{r}
ggplot(data=rainfall_g8_up[rainfall_g8_up["location"] == "loc_1", ], aes(x=weekday, y=value, color=weekday)) + 
  geom_boxplot()  +
  labs(title="Distribution of total of rainfall in each weekday" , y="Rainfall", x="Years", legend="Location") +    theme(plot.title=element_text(color="black", size=16, face="bold.italic", hjust="0.5"))
```

```{r}
rainfall_g8_up$month_c <- factor(rainfall_g8_up$month_c, levels=as.character(1:12))

ggplot(data=rainfall_g8_up[rainfall_g8_up["location"] == "loc_1", ], aes(x=month_c, y=value, color=month_c)) + 
  geom_boxplot()  +
  labs(title="Distribution of total of rainfall in each location" , y="Rainfall", x="Years", color="Month") +    theme(plot.title=element_text(color="black", size=16, face="bold.italic", hjust="0.5")) +
  scale_fill_manual(breaks=as.character(1:12))
```

```{r}
ggplot(data=rainfall_g8_up, aes(x=value)) + 
  geom_histogram() +
  facet_wrap(.~location) + 
  labs(title="Distribution of total of rainfall in each location" , y="Rainfall", x="Years", color="Month") +   theme(plot.title=element_text(color="black", size=16, face="bold.italic", hjust="0.5")) +
  scale_fill_manual(breaks=as.character(1:12))
```

```{r}
ggplot(data=rainfall_g8_up[rainfall_g8_up["value"] > 25, ], aes(x=value)) + 
  geom_histogram() +
  facet_wrap(.~location) + 
  labs(title="Distribution of total of rainfall above 25 in each location" , y="Rainfall", x="Years", color="Month") +   theme(plot.title=element_text(color="black", size=16, face="bold.italic", hjust="0.5")) +
  scale_fill_manual(breaks=as.character(1:12))
```

We can see here that in our data we've got a lot of extreme values. Usually it is not raining, but from time to time the rainfall is really strong.
In each station we can observe a similar distribution of data.

For the above reasons, we can conclude that the data meet the assumptions of Extreme Value Theory.

### Fit to GEV distribution

```{r}
x1 <- rainfall_g8_gyear_month[rainfall_g8_gyear_month["location"] == "loc_1", ]$max_rainfall
x2 <- rainfall_g8_gyear_month[rainfall_g8_gyear_month["location"] == "loc_2", ]$max_rainfall
x3 <- rainfall_g8_gyear_month[rainfall_g8_gyear_month["location"] == "loc_3", ]$max_rainfall
x4 <- rainfall_g8_gyear_month[rainfall_g8_gyear_month["location"] == "loc_4", ]$max_rainfall
x5 <- rainfall_g8_gyear_month[rainfall_g8_gyear_month["location"] == "loc_5", ]$max_rainfall

f1 <- gev.fit(x1)
f2 <- gev.fit(x2)
f3 <- gev.fit(x3)
f4 <- gev.fit(x4)
f5 <- gev.fit(x5)
```

```{r}
gev.diag(f1)
```

```{r}
print(paste("Shape:", as.character(f1$mle[3]), sep= " "))
```
Shape parameter is greater than zero, so this distribution is heavy tailed and its commonly known as Fréchet distribution.

The distribution appears to be a good fit to the theoretical data (qq plot analysis).

```{r}
gev.diag(f2)
```

```{r}
print(paste("Shape:", as.character(f2$mle[3]), sep= " "))
```

Shape parameter is greater than zero, so this distribution is heavy tailed and its commonly known as Fréchet distribution.

The distribution appears to be a good fit to the theoretical data (qq plot analysis).

```{r}
gev.diag(f3)
```

```{r}
print(paste("Shape:", as.character(f3$mle[3]), sep= " "))
```

Shape parameter is greater than zero, so this distribution is heavy tailed and its commonly known as Fréchet distribution.

In larger quantiles, the distribution seems to diverge from the theoretical values. In other quantiles it is ok.

```{r}
gev.diag(f4)
```

```{r}
print(paste("Shape:", as.character(f4$mle[3]), sep= " "))
```

Shape parameter is greater than zero, so this distribution is heavy tailed and its commonly known as Fréchet distribution.

The distribution appears to be a good fit to the theoretical data (qq plot analysis).


```{r}
gev.diag(f5)
```

```{r}
print(paste("Shape:", as.character(f5$mle[3]), sep= " "))
```

Shape parameter is greater than zero, so this distribution is heavy tailed and its commonly known as Fréchet distribution.

The distribution appears to be a good fit to the theoretical data (qq plot analysis).


### Choosing threshold

```{r Fig1, echo=TRUE, fig.height=12, fig.width=5}

par(mfrow=c(5, 1))
mrl.plot(rainfall_g8$loc_1) # 30, 60
mrl.plot(rainfall_g8$loc_2) # 20, 40
mrl.plot(rainfall_g8$loc_3) # 40, 60
mrl.plot(rainfall_g8$loc_4) # 50
mrl.plot(rainfall_g8$loc_5) # 40
```
```{r}
gpd.fitrange(rainfall_g8$loc_1, 0, 50)
```
```{r}
gpd.fitrange(rainfall_g8$loc_2, 0, 50)
```

```{r}
gpd.fitrange(rainfall_g8$loc_3, 0, 50)
```

```{r}
gpd.fitrange(rainfall_g8$loc_4, 0, 50)
```

```{r}
gpd.fitrange(rainfall_g8$loc_5, 0, 50)
```

We decided to choose following thresholds due to parameter stability:
loc1 - 25
loc2 - 25
loc3 - 25
loc4 - 25
loc5 - 25

### Fit data to General Pareto Distribution

```{r}
f1_threshold <- gpd.fit(rainfall_g8$loc_1, 25)
f2_threshold <- gpd.fit(rainfall_g8$loc_2, 25)
f3_threshold <- gpd.fit(rainfall_g8$loc_3, 25)
f4_threshold <- gpd.fit(rainfall_g8$loc_4, 25)
f5_threshold <- gpd.fit(rainfall_g8$loc_5, 25)
```

nexc - here we can see the number of observations above threshold
convs - looking at the values, we can conclude that all convergences were successful
nllh - negative logarithm of likelihood at maximum
mle - proportion of data above threshold
rate - standard errors

### Confidence interval for shape parameters

```{r}
gpd.profxi(f1_threshold, -0.2, 0.5)
```

```{r}
gpd.profxi(f2_threshold, -0.2, 0.7)
```

```{r}
gpd.profxi(f3_threshold, -0.2, 1.0)
```

```{r}
gpd.profxi(f4_threshold, -0.2, 0.5)
```

```{r}
gpd.profxi(f5_threshold, -0.2, 1.0)
```
We can see that the shape parameter almost in every case takes positive values. 
Thus, we are dealing here with a distributions with a heavy tail

### Confidence interval of return level for return period=100 in each weather station

```{r}
gpd.prof(f1_threshold, m=100, 60, 200)
```
```{r}
gpd.prof(f2_threshold, m=100, 60, 250)
```
```{r}
gpd.prof(f3_threshold, m=100, 60, 500)
```
```{r}
gpd.prof(f4_threshold, m=100, 60, 300)
```
```{r}
gpd.prof(f5_threshold, m=100, 60, 300)
```

We can observe similar confidence intervals for stations 2, 5. 
For station number 3 the values are much higher, while for stations number 1 and 4 they are slightly lower

### Extremal index for first weather station for threshold equals to 25

```{r}
theta <- extremalindex(rainfall_g8$loc_1, method="runs", threshold=25)
theta
```

Extremal index value of 0.96 means that the Gumbel distribution for the stationary process is almost 
identical to that formed by independent random variables with the same marginal distributions.
It measures the clustering of data. Values close to one mean that the maximum values do not tend to cluster.


```{r}
theta <- extremalindex(rainfall_g8$loc_2, method="runs", threshold=25)
theta
```
Extremal index value of 0.96 means that the Gumbel distribution for the stationary process is almost 
identical to that formed by independent random variables with the same marginal distributions.
It measures the clustering of data. Values close to one mean that the maximum values do not tend to cluster.

```{r}
theta <- extremalindex(rainfall_g8$loc_3, method="runs", threshold=25)
theta
```
Extremal index value of 0.95 means that the Gumbel distribution for the stationary process is almost 
identical to that formed by independent random variables with the same marginal distributions.
It measures the clustering of data. Values close to one mean that the maximum values do not tend to cluster.
```{r}
theta <- extremalindex(rainfall_g8$loc_4, method="runs", threshold=25)
theta
```
Extremal index value of 0.95 means that the Gumbel distribution for the stationary process is almost 
identical to that formed by independent random variables with the same marginal distributions.
It measures the clustering of data. Values close to one mean that the maximum values do not tend to cluster.
```{r}
theta <- extremalindex(rainfall_g8$loc_5, method="runs", threshold=25)
theta
```
Extremal index value of 0.90 means that the Gumbel distribution for the stationary process is almost 
identical to that formed by independent random variables with the same marginal distributions.
It measures the clustering of data. Values close to one mean that the maximum values do not tend to cluster.


```{r}
# Function to estimate the empirical extremogram between two locations
computeExtremalCoeff <- function(pos1, pos2, database, u) {
# Subset the data above the threshold value
  subdata <- database[database[, pos1] > u & !is.na(database[, pos1]) &
                      database[, pos2] > u & !is.na(database[, pos2]), ]
  # Compute the extremal coefficient using the subsetted data
  theta <- nrow(subdata) / length(which((database[, pos1] > u & !is.na(database[, pos1]))))
  return(theta)
}
```

```{r}
quantiles
```

Multivariate extremal coefficient
```{r}

##### CALCULATION OF EXTREMAL COEFFICIENT FOR ALL THE PAIRS ###########
#### get them in separate dataframe
loc1_df <- subset(rainfall_g8_gyear_month,location == 'loc_1')
loc2_df <- subset(rainfall_g8_gyear_month,location == 'loc_2')
loc3_df <- subset(rainfall_g8_gyear_month,location == 'loc_1')
loc4_df <- subset(rainfall_g8_gyear_month,location == 'loc_2')
loc5_df <- subset(rainfall_g8_gyear_month,location == 'loc_2')

threshold <- 25 ## according to what was calculated, for the location separately


combined_df <- cbind(loc1_df["max_rainfall"], loc2_df["max_rainfall"],loc3_df["max_rainfall"],loc4_df["max_rainfall"],loc5_df["max_rainfall"])
### 

# local quantiles for each variable
quantiles <- apply(combined_df[,1:5], 2, quantile, probs = 0.95, na.rm = TRUE)

# the number of variables
n_vars <- ncol(combined_df[,1:5])
## compute the extreme coefficient for all the pairs

# matrix to store the extremal coefficients
theta <- matrix(0, nrow = n_vars, ncol = n_vars)

for (i in 1:(n_vars - 1)) {
  for (j in (i + 1):n_vars) {
    theta[i, j] <- computeExtremalCoeff(pos1 = i, pos2 = j, database = combined_df[,1:5], threshold)
    theta[j, i] <- theta[i, j] # Store the symmetric value
  }
}

#The estimated extremal coefficients
colnames(theta) <- rownames(theta) <- names(combined_df[,1:5])
print(theta)

```

####FIRST AND SECOND LOCATION###

```{r}

# Set quantile level

quantileLevel <- 0.95
quantile_1 <- quantile(combined_df[, 1], quantileLevel, na.rm = TRUE)
quantile_2 <- quantile(combined_df[, 2], quantileLevel, na.rm = TRUE)

# data above and below threshold quantile
tmp_below_1 <- combined_df[combined_df[, 1] <= 25, 1]
tmp_above_1 <- combined_df[combined_df[, 1] > 25, 1] 
tmp_below_2 <- combined_df[combined_df[, 2] <= 25, 2]
tmp_above_2 <- combined_df[combined_df[, 2] > 25, 2] 

scale_1 <- 7.3655429
scale_2 <- 6.8489896
shape_1 <- 0.1405578
shape_2 <- 0.1559929

#### first location

empiricalCdf_1 <- ecdf(combined_df[, 1])

empiricalCdf_2 <- ecdf(combined_df[, 2])

normalized_database <- matrix(NA, nrow = length(combined_df[, 1]), ncol = 2)

normalized_database[combined_df[, 1] <= quantile_1, 1] <- 1 / (1 - empiricalCdf_1(combined_df[combined_df[, 1] <= quantile_1 ,1])) - 1

normalized_database[combined_df[, 1] > quantile_1, 1] <- 1 / ((1 - quantileLevel) * (1 + shape_1*(combined_df[combined_df[, 1] > quantile_1, 1] - quantile_1) / scale_1) ^ (-1 / shape_1)) - 1


normalized_database[combined_df[, 2] <= quantile_1, 2] <- 1 / (1 - empiricalCdf_2(combined_df[combined_df[, 2] <= quantile_1 ,2])) - 1

normalized_database[combined_df[, 2] > quantile_2, 2] <- 1 / ((1 - quantileLevel) * (1 + shape_2 * (combined_df[combined_df[, 2] > quantile_2, 2] - quantile_2) / scale_2) ^ (-1 / shape_2)) - 1

```
CALCULATION OF COPULA 
```{r}
ecdf_1 <- ecdf(normalized_database[, 1])
ecdf_2 <- ecdf(normalized_database[, 2])
ecdf_loc_1 <- ecdf_1(normalized_database[, 1])
ecdf_loc_2 <- ecdf_1(normalized_database[, 2])
edfc =rbind(ecdf_loc_1,ecdf_loc_2)
# Fit a copula with Frechet margins, gumbel copula (for extremne values)
fit <- BiCopSelect(ecdf_loc_1, ecdf_loc_2)


### pickands function


# Calculate the empirical exceedance probability for each observation in each variable
u_1 <- sapply(normalized_database[, 1], function(x) ecdf_1(x))
u_2 <- sapply(normalized_database[, 2], function(x) ecdf_2(x))

# Combine the transformed variables into a matrix
u <- cbind(u_1, u_2)

n <- nrow(u)
k <- 30  # number of intervals for estimating the Pickands function
U <- seq(1/k, 1, length = k)  # partition [0, 1] into k intervals
n_i <- rep(0, k)
S_i <- rep(0, k)

for (i in 1:n) {
  j <- sum(U <= u[i, 1])
  n_i[j] <- n_i[j] + 1
  S_i[j] <- S_i[j] + max(u[i,]) - u[i, 1]
}

# Estimating the Pickands function
n_i <- n_i[n_i > 0]  # remove empty intervals
S_i <- S_i[n_i > 0]
mu_i <- S_i/n_i
theta_hat <- log(mu_i[-1]/mu_i[-length(mu_i)]) / log(U[-1]/U[-length(U)])

```

DETERMINING THE EXTREMAL COEFFICIENT
```{r}
#### bivariate EXTREMAL COEFFICIENT
theta <- 1 - 2 * BiCopTau(fit, c(0, 0))
## theta is the coeeficinet
```


